---
title: "Practice: R and Tidyverse for Clinical Data"
output: html_document
---

<!--
Dataset: MIMIC-III Clinical Database Demo v1.4 (real, openly-accessible, 100 patients)
See ../00_setup_mimic_demo.md for how to download the CSVs.
All table/column names below are verified against the real MIMIC-III schema (same source used in the
SQL practice file, 3_Tools_SQL/practice/01_clinical_queries.sql).
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(readr)
library(lubridate)

# Adjust this path to wherever you downloaded/unzipped the MIMIC-III demo CSVs
data_dir <- "mimic-iii-demo"
```

## 1. Load the tables

```{r load-data}
patients        <- read_csv(file.path(data_dir, "PATIENTS.csv"))
admissions      <- read_csv(file.path(data_dir, "ADMISSIONS.csv"))
icustays        <- read_csv(file.path(data_dir, "ICUSTAYS.csv"))
diagnoses_icd   <- read_csv(file.path(data_dir, "DIAGNOSES_ICD.csv"))
d_icd_diagnoses <- read_csv(file.path(data_dir, "D_ICD_DIAGNOSES.csv"))
labevents       <- read_csv(file.path(data_dir, "LABEVENTS.csv"))
d_labitems      <- read_csv(file.path(data_dir, "D_LABITEMS.csv"))
```

## 2. Orientation

```{r orientation}
nrow(patients)
n_distinct(admissions$subject_id)
```

## 3. filter() — patients who died (equivalent to SQL's WHERE)

```{r filter-deceased}
deceased <- patients |>
  filter(expire_flag == 1) |>
  select(subject_id, gender, dob, dod)

deceased
```

## 4. left_join() / inner_join() — diagnoses with readable labels

This is the tidyverse equivalent of the SQL JOIN between `DIAGNOSES_ICD` (fact table) and
`D_ICD_DIAGNOSES` (lookup table):

```{r join-diagnoses}
diagnoses_labeled <- diagnoses_icd |>
  inner_join(d_icd_diagnoses, by = "icd9_code") |>
  arrange(subject_id, seq_num)

diagnoses_labeled |> head(20)
```

## 5. group_by() + summarise() — most common diagnoses (equivalent to SQL GROUP BY)

```{r group-summarise}
top_diagnoses <- diagnoses_labeled |>
  distinct(subject_id, hadm_id, short_title) |>   # avoid double-counting within an admission
  count(short_title, name = "n_admissions") |>
  arrange(desc(n_admissions))

top_diagnoses |> head(10)
```

## 6. Building an analytic cohort table (multi-table join, mirrors the SQL 3-table join)

```{r cohort-table}
primary_diagnosis <- diagnoses_icd |>
  filter(seq_num == 1) |>          # seq_num = 1 -> primary diagnosis
  inner_join(d_icd_diagnoses, by = "icd9_code") |>
  select(subject_id, hadm_id, primary_diagnosis = short_title)

cohort <- patients |>
  select(subject_id, gender) |>
  inner_join(admissions |> select(subject_id, hadm_id, admittime, admission_type),
             by = "subject_id") |>
  inner_join(primary_diagnosis, by = c("subject_id", "hadm_id")) |>
  arrange(admittime)

cohort |> head(10)
```

## 7. Working with lab results

```{r lab-results}
lab_results_labeled <- labevents |>
  inner_join(d_labitems, by = "itemid")

# Most frequently measured labs
lab_results_labeled |>
  count(label, fluid, category, name = "n_results") |>
  arrange(desc(n_results)) |>
  head(10)

# Abnormal results only
abnormal_labs <- lab_results_labeled |>
  filter(flag == "abnormal") |>
  select(subject_id, label, charttime, value, valueuom, flag) |>
  arrange(charttime)

abnormal_labs |> head(20)
```

## 8. Dates: ICU length of stay

`ICUSTAYS.los` is precomputed by MIMIC-III in days — no manual date arithmetic needed here,
but this is the natural place to introduce `lubridate` for later modules that don't have a
precomputed duration column.

```{r icu-los}
icustays |>
  select(subject_id, hadm_id, icustay_id, first_careunit, intime, outtime, los) |>
  arrange(desc(los)) |>
  head(10)
```

## 9. Visualization with ggplot2

```{r plot-diagnoses}
top_diagnoses |>
  head(10) |>
  ggplot(aes(x = reorder(short_title, n_admissions), y = n_admissions)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 primary diagnoses by number of admissions",
    subtitle = "MIMIC-III Clinical Database Demo (n = 100 patients)",
    x = NULL,
    y = "Number of admissions"
  )
```

```{r plot-los}
icustays |>
  ggplot(aes(x = los)) +
  geom_histogram(binwidth = 1) +
  labs(
    title = "Distribution of ICU length of stay",
    subtitle = "MIMIC-III Clinical Database Demo (n = 100 patients)",
    x = "Length of stay (days)",
    y = "Count"
  )
```
